### 2025-11-09 Proposals/2025-NSF-EntropicGeometry/NarrativeBody.tex
Fix typos, add refs, remove blue text

While the structures of the state spaces of classical and quantum mechanics are different, their spaces of ensembles have a rather similar structure: they both consist of a bounded convex subset of a metrizable and second-countable locally convex topological vector space, where the metric can be derived from the entropy through the square root of the Jensen-Shannon divergence $JSD = S\left(\frac{1}{2} \ens[a] + \frac{1}{2} \ens[b]\right) - \left(\frac{1}{2}S(\ens[a]) + \frac{1}{2}S(\ens[b])\right)$~\cite{Jensen-ShannonOriginal,HayashiQuantumInfo,majtey2005jensen}. The JSD is, in both cases, the square of a distance function and can be interpreted as how much the average entropy increases under an equal mixture of two ensembles. There are, therefore, three common structures at play: a topology (which physically represents experimental verifiability), a convex structure (which physically represents the ability to make statistical mixtures) and an entropy function (which physically represents the variability of the elements within each ensemble). While the axioms for the topology and for convex spaces(CITE) are readily available, to our knowledge, the correct axioms for an entropy function and its required compatibility condition with the other two structure still need to be identified.

@@

The central focus is the intuitive notion of statistical ensemble (i.e.~the collection of all outputs of a preparation procedure), as this seems to be particularly suitable for generalization for a number of both practical and conceptual considerations. In terms of the mathematics, as we have seen, the spaces of statistical ensembles of classical and quantum mechanics can be made to look very similar. From the experimental side, statistical ensembles are what we produce and measure experimentally. On the conceptual side, the reproducibility requirement of scientific investigation means that physical laws are of the form ``whenever I prepare a system according to this procedure, I will find this range of outcomes.'' That is, physical laws are relationships between ensembles. Therefore, it makes sense to require that any physical theory must \emph{at least} posit a space of ensembles. This provides us with a strong conceptual starting point, and properties of ensembles are what need to be specified through axioms. We will therefore call \textbf{ensemble space} a set endowed with the suitable topology, convex structure and entropic structure that generalizes the spaces of classical and quantum ensembles.

Before proceeding, we should not that the entropic structure is not crucial for the geometry of the ensemble space, but it is also intimately related to the geometry of the state space of both classical and quantum mechanics. One direction we want to pursue, in fact, is to recover to the geometric structure of the state space from it entropic structure.

For example, as we saw in classical mechanics, the symplectic form gives us the Liouville measure which in turn is used to define the entropy. Therefore, the entropy of an ensemble is defined from the geometry of the state space. However, note that the entropy for a uniform distribution $\rho_U$ over a set $U$ is given by $S(\rho_U) = \log \mu(U)$. Therefore, if we are given the entropy for all probability measures, we can recover the Liouville measure by setting $\mu(U) = 2^{S(\rho_U)}$. While this does not recover the symplectic form directly, note that the symplectic form can also be used to define the entropy of the marginal distributions along each degree of freedom. Around a point, in fact, we can define Darboux coordinates and define joint probability distributions that are the product of independent marginals along each pair of position and momentum. One finds a similar relationship between the entropy of uniform marginal distributions and the integral of the symplectic form over the support of the marginal. From these, we believe we can recover the symplectic form.

In quantum mechanics, the geometry is given by the Born rule
\begin{equation}
	p(\phi|\psi) = \frac{\< \phi | \psi \>\< \psi | \phi \>}{\< \psi | \psi \>\< \phi | \phi \>}.
\end{equation}
as this is the cosine square of the distance function of the Fubini-Study metric. Given two states $\psi$ and $\phi$, the ensemble given by their equal mixture, $\rho_{\psi\phi} = \frac{1}{2}|\psi\>\<\psi|+\frac{1}{2}|\phi\>\<\phi|$,  has von Neumann entropy $S(\rho_{\psi\phi}) = I\left(\frac{1 + \sqrt{p(\phi|\psi)}}{2} \right)$ where $I$, as before, is the Shannon entropy for two elements. As before, the entropy is defined in terms of the inner product. However, the relationship between $S(\rho_{\psi\phi})$ and $p(\phi|\psi)$ is invertible. This means that, given the entropy of the equal mixtures of all pairs of states we can reconstruct the Born rule and, therefore, the geometry.

Lastly, as port of our larger program, we have shown that classical mechanics emerges as the high entropy limit of quantum mechanics. These insights show that the entropic structure is more fundamental that one may expect, and that finding its axiomatic characterization should allow for a more precise generalization of a series of results. We stress that the axioms that characterize the entropic structure of physical theories should be minimal, in the sense that, in the spirit of analogous work on the foundations of mathematics, like set theory or reverse mathematics~\cite{RevMathBook}, the axioms can be shown to be satisfied by the different theories (which now become models of the larger theory) but also allow one to rederive the entropy formula for each specific theory. Part of the work, then, is to understand the implication relationships between different requirements. Lastly, note that the Lie algebras of both classical and quantum mechanics, Poisson brackets and commutators, respectively, have a special relationship with the geometric and entropic structures, which may provide us with a template to generalize those structures as well.

Before delving into the axiomatic, note that in all above three cases the definition of the entropy is tightly bound to the definition of the space. Therefore, we need to understand how one can give a single characterization that is general enough to encompass the three cases, while constrained enough to recover the correct expression in each case. Fortunately, in our preliminary work we have found the right approach. In all cases, the entropy possesses an upper bound to its increase under convex combination: $S(p\ens[a] + \bar{p} \ens[b]) \leq I(p) + p S(\ens[a]) + \bar{p} S(\ens[b])$ where $\bar{p} = 1-p$ and $I(p) = -p \log p - \bar{p}\log \bar{p}$. This maximum increase is given, in the classical case, when the two measures have disjoint support; in the quantum case, it is given when the eigenspaces of non-zero eigenvalues are orthogonal. Our approach, then, is to proceed in the opposite way: impose a generic entropy function to have an upper bound and enough properties to show that said upper bound must correspond to the Shannon entropy formula~\cite{Shannon}. Since both classical and quantum mechanics ensembles can always be decomposed into orthogonal components, the different entropy formulas can be recovered in each case as a recursive expansion of the upper bound.

@@

The second axiom imposes a convex structure, which represents the ability to perform statistical mixtures. Given two ensembles $\ens[a], \ens[b] \in \Ens$, we can always choose a probability $p \in [0,1]$ and select $\ens[a]$ $p$ percent of the times and $\ens[b]$ $\bar{p} = 1-p$ percent of the times. This gives the ensemble space a convex structure as $p \ens[a] + \bar{p} \ens[b] \in \Ens$ for all $\ens[a], \ens[b] \in \Ens$ and $p \in [0,1]$. This insight is already used in the literature, particularly in Generalized Probabilistic Theories (GPTs); see e.g.~Refs.~\cite{plavala2023,wilce2025generalizedprobabilitytheorynotes,DArianoBook}. The main innovation in this respect is that we do not limit ourselves to finite combinations and we do not simply allow all infinite combinations (which can be shown to include physically untenable cases~\cite{baccetti2013infinite,aop-unphysHilbert}). Rather, we let the topology do the work. That is, infinite convex combinations exist only if they are the topological limit of finite convex combinations. On physical grounds, the mixing operation must be continuous in the natural topology defined by experimental verifiability, much like addition and scalar multiplication are continuous in a topological vector space. Given the analogy, we call \textbf{topological convex space} a convex space that is also a topological space for which the mixing operation is continuous.

Topological convex spaces have not been developed in the literature, and this structure alone presents interesting sets of questions.  For example, it is established that a convex space embeds into a vector space if and only if it is cancellative, that is, if $p \ens[a] + \bar{p} \ens = p \ens[b] + \bar{p} \ens$ for some $p \in (0,1)$ implies $\ens[a] = \ens[b]$. The question now is what are necessary and sufficient conditions for a topological convex space to embed continuously into a topological vector space? Can we have, for example, a Hausdorff second-countable topological convex space that is not a continuous embedding of a topological vector space? Topological vector spaces present a number of nice results which may or may not hold in topological convex spaces. For example, any $\mathsf{T}_1$ second-countable topological vector space is metrizable. Does that same result hold for topological convex spaces? Another interesting aspect is the notion of closure. In a topological vector space, the topological closure of a convex set is convex. We found that this is true in topological convex spaces as well. Additionally, there is another type of closure that may be more interesting physically. We can define the $\sigma$-hull of a set $A$ as the set of possible infinite convex combinations, which can also be shown to be a convex set. This will be a superset of the hull of $A$, because it will include all finite convex combinations, and it will be a subset of the closed hull of $A$, as all infinite convex combinations are topological limits. Moreover, it can be shown that, in some cases, the hull, the $\sigma$-hull and the closed hull are all different. From the physics side, the $\sigma$-hull may be a more physically meaningful closure as it represents the true limit process of statistical mixing, the operation the axiom is characterizing. The subtle issue is that convex combinations always return elements ``inside'' the set, while linear combinations can also return the ``edges.'' Therefore, there are already interesting topics to study even without imposing the entropic structure.

@@

Though we do not believe the current version of the axiom lists all necessary properties of the entropic structure, as discussed below, it does list enough of them to already prove interesting results. For example, we can show that $I(p)$, the maximum entropy increase during mixing, must be, up to a multiplicative constant, the Shannon entropy~\cite{Shannon} $I(p)= - p \log p - \bar{p} \log \bar{p}$. The proof is general in the sense that it does not assume whether the specific model is classical, quantum or something else entirely. Moreover, in the case of both classical and quantum mechanics, states can be decomposed into orthogonal components, which means the upper variability is an equality and can be used to calculate the entropy from that decomposition. This will allows us to recover the expressions for the entropy in the three different cases, classical discete, classical continuous, and quantum. Similarly, we will be able to show that, from this structure alone, the geometric structure of both classical and quantum mechanics can be recovered.

Our preliminary work also shows that an entropy satisfying those properties forces the convex space to be cancellative, meaning that it embeds into a vector space. That is, a convex space that does not embed into a vector space cannot posses an entropic structure. We can, in fact, produce a model of that case and see exactly why the entropic structure fails. We can also show that an ensemble space must be ``directionally bounded,'' meaning its intersection with any affine line is a finite interval. It is an open question as to whether the ensemble space must be a bounded set. We believe we can also show that some examples of Generalized Probabilistic Theories are inconsistent with an entropic structure and therefore cannot represent a physically meaningful space of ensembles. In other words, the entropic structure constrains the types of spaces possible for physical theories in surprising and interesting ways.

@@

Given an internal point $\ens[a] \in \Ens$ (i.e. for every $\ens[b] \in \Ens$ we can find $\ens[c] \in \Ens$ such that $\ens[a] = p \ens[b] + \bar{p} \ens[c]$ from some $p \in (0,1)$) we can construct an embedding vector space $V_{\ens[a]}$. If $V_{\ens[a]}$ is finite dimensional, we can show that $V_{\ens[a]}$ is a neighborhood of $\ens[a]$, which means that the interior of $\Ens$ is a manifold. If the entropy is twice differentiable with respect to the convex coefficients, the strict concavity of the entropy means that the Hessian is negative definite, which allows us to define a metric tensor. On a linear coordinate chart we can define $g_{\ens}( v, w ) = - \frac{\partial^2 S}{\partial \ens ^2}( v, w )$ where $v$ and $w$ are elements of the tangent space at $\ens \in \Ens \subset V_{\ens[a]}$. Since all Hessians are symmetric, $g_{\ens}$ will be bilinear, symmetric and positive definite: it is a metric tensor and $V_{\ens[a]}$ is a Hessian manifold~\cite{hessianmanifoldsShima1997,hessianmanifoldsliu2025}. In classical and quantum mechanics it recovers the Fisher information metric (see e.g.~Refs.~\cite{facchi2010,AmariInfoGeom}). The immediate goal, then, is to see how much of information geometry can be generalized to this case. If a locally convex topology can be established, the further goal is to extend this to the infinite-dimensional case.

Another construction allows us to generalize the notion of count of states given, in classical mechanics, by the Liouville measure. This can be done by noting two things: for a uniform distribution with support $U \subseteq X$, $\mu(U) = 2^{S(\rho_U)}$, and the uniform distribution over $U$ has the highest entropy of all distributions with support $U$. Given a set $A \subseteq \Ens$ of ensembles, we can then define the state capacity as the supremum of the exponential of the entropy within the hull of $A$, which physically means the highest entropy achievable with a statistical mixture of elements of $A$. This set function has nice properties: it is non-negative, monotonic, subadditive in general and additive on orthogonal sets of ensembles. Note that if $A_U \subseteq \Ens$ is the subspace of all probability measures with support $U \subseteq X$, the state capacity of $A_U$ recovers the Liouville measure of $U$. Moreover, $A_U$ is orthogonal to $A_V$ if and only if $U$ and $V$ are disjoint. This means that the state capacity restricted to the lattice of those subspaces is an additive measure. Interestingly enough, one can follow the same construction in the quantum case, and the measure will return the dimensionality of the subspace. Concerning non-additive measures, there are additional interesting questions as one can create a non-additive version of probability measures, which may point to a non-additive version of probability theory. We have already looked at this problem, and, unfortunately, the standard techniques that have been developed in that area are not sufficient to handle the cases of interest to us within physical theories.

@@

\textcolor{blue}{The immediate significance of the proposed research stems from the realization that, though classical information theory and quantum information theory are presented as two separate subjects in textbooks, their numerous corresponding features make them an interesting target for generalization.  The broader significance lies in the overall approach of the research as well as its specific potential to provide a common axiomatic foundation for \emph{all} physical theories.  Currently within physics, we typically do not know from where the various mathematical structures emerge, making it unclear in which directions they can be generalized in order to develop new theories that are physically meaningful.  Our approach instead seeks to \emph{find axioms from physical requirements}, i.e.~the axiomatics are justified from the physics.  Such an axiomatic foundation common to all physical theories would serve not only to better ground current theories and elucidate their relationships, but also to constrain potential future physical theories.} While there exists other work attempting to reconstruct quantum mechanics from physical principles, for example in Refs.~\cite{hardy2001quantumaxioms,caticha2008,dakic2009,Chiribella2010,Masanes_2011}, we are not aware of any literature attempting to relate the geometric and entropic structures of either classical or quantum mechanics, much less those of a general physical theory. Works trying to reconstruct quantum mechancis from physical premises are typically in finite dimensions, do not generalize to classical mechanics, and do not provide a mathematically rigorous path to generalize to classical or quantum field theories.

@@

\noindent
\textbf{Relationship to other current funding:} PI Aidala's current funding from the NSF Division of Physics is for work on LHCb experiment at CERN, and from the NSF AccelNet program is to increase collaboration on quantum chromodynamics (QCD) research between the U.S.~and Latin America.  The PI's funding from the Department of Energy is for work at the Relativistic Heavy Ion Collider (RHIC) on the PHENIX and sPHENIX experiments.  These grants are for a completely separate line of research. Her grant from the John Templeton Foundation, titled ``Why classical probability and classical information theory are incompatible with quantum mechanics and quantum contextuality,'' supports the development of novel non-additive measures (set functions) suitable for quantifying the number of states in quantum mechanics~\cite{aop-nonaddmeas}.  There is no direct overlap with the activities proposed here.  Co-PI Uribe does not have any current funding.

>>>>

While the structures of the state spaces of classical and quantum mechanics are different, their spaces of ensembles have a rather similar structure: they both consist of a bounded convex subset of a metrizable and second-countable locally convex topological vector space, where the metric can be derived from the entropy through the square root of the Jensen-Shannon divergence $JSD = S\left(\frac{1}{2} \ens[a] + \frac{1}{2} \ens[b]\right) - \left(\frac{1}{2}S(\ens[a]) + \frac{1}{2}S(\ens[b])\right)$~\cite{Jensen-ShannonOriginal,HayashiQuantumInfo,majtey2005jensen}. The JSD is, in both cases, the square of a distance function and can be interpreted as how much the average entropy increases under an equal mixture of two ensembles. There are, therefore, three common structures at play: a topology (which physically represents experimental verifiability), a convex structure (which physically represents the ability to make statistical mixtures) and an entropy function (which physically represents the variability of the elements within each ensemble). While the axioms for the topology and for convex spaces~\cite{Gudder1973,fritz2015convexspaces} are readily available, to our knowledge, the correct axioms for an entropy function and its required compatibility condition with the other two structures still need to be identified.

@@

The central focus is the intuitive notion of statistical ensemble (i.e.~the collection of all outputs of a preparation procedure), as this seems to be particularly suitable for generalization for a number of both practical and conceptual considerations. In terms of the mathematics, as we have seen, the spaces of statistical ensembles of classical and quantum mechanics can be made to look very similar. From the experimental side, statistical ensembles are what we produce and measure experimentally. On the conceptual side, the reproducibility requirement of scientific investigation means that physical laws are of the form, ``Whenever I prepare a system according to this procedure, I will find this range of outcomes.'' That is, physical laws are relationships between ensembles. Therefore, it makes sense to require that any physical theory must \emph{at least} posit a space of ensembles. This provides us with a strong conceptual starting point, and properties of ensembles are what need to be specified through axioms. We will therefore call \textbf{ensemble space} a set endowed with the suitable topology, convex structure and entropic structure that generalizes the spaces of classical and quantum ensembles.

Before proceeding, we should note that the entropic structure is not only crucial for the geometry of the ensemble space, but it is also intimately related to the geometry of the state space of both classical and quantum mechanics. One direction we want to pursue, in fact, is to recover the geometric structure of the state space from its entropic structure.

For example, as we saw in classical mechanics, the symplectic form gives us the Liouville measure, which in turn is used to define the entropy. Therefore, the entropy of an ensemble is defined from the geometry of the state space. However, note that the entropy for a uniform distribution $\rho_U$ over a set $U$ is given by $S(\rho_U) = \log \mu(U)$. Therefore, if we are given the entropy for all probability measures, we can recover the Liouville measure by setting $\mu(U) = 2^{S(\rho_U)}$. While this does not recover the symplectic form directly, note that the symplectic form can also be used to define the entropy of the marginal distributions along each degree of freedom. Around a point, in fact, we can define Darboux coordinates and define joint probability distributions that are the product of independent marginals along each pair of position and momentum. One finds a similar relationship between the entropy of uniform marginal distributions and the integral of the symplectic form over the support of the marginal. From these, we believe we can recover the symplectic form.

In quantum mechanics, the geometry is given by the Born rule
\begin{equation}
	p(\phi|\psi) = \frac{\< \phi | \psi \>\< \psi | \phi \>}{\< \psi | \psi \>\< \phi | \phi \>},
\end{equation}
as this is the cosine squared of the distance function of the Fubini-Study metric. Given two states $\psi$ and $\phi$, the ensemble given by their equal mixture, $\rho_{\psi\phi} = \frac{1}{2}|\psi\>\<\psi|+\frac{1}{2}|\phi\>\<\phi|$,  has von Neumann entropy $S(\rho_{\psi\phi}) = I\left(\frac{1 + \sqrt{p(\phi|\psi)}}{2} \right)$ where $I$, as before, is the Shannon entropy for two elements. As before, the entropy is defined in terms of the inner product. However, the relationship between $S(\rho_{\psi\phi})$ and $p(\phi|\psi)$ is invertible. This means that, given the entropy of the equal mixtures of all pairs of states we can reconstruct the Born rule and, therefore, the geometry.

Lastly, as part of our larger program, we have shown that classical mechanics emerges as the high-entropy limit of quantum mechanics~\cite{aop-classicallimit}. These insights show that the entropic structure is more fundamental than one may expect, and that finding its axiomatic characterization should allow for a more precise generalization of a series of results. We stress that the axioms that characterize the entropic structure of physical theories should be minimal, in the sense that, in the spirit of analogous work on the foundations of mathematics, like set theory or reverse mathematics~\cite{RevMathBook}, the axioms can be shown to be satisfied by the different theories (which now become models of the larger theory) but also allow one to rederive the entropy formula for each specific theory. Part of the work, then, is to understand the implication relationships between different requirements. Lastly, note that the Lie algebras of both classical and quantum mechanics, Poisson brackets and commutators, respectively, have a special relationship with the geometric and entropic structures, which may provide us with a template to generalize those structures as well.

Before delving into the axiomatics, note that in all three cases above the definition of the entropy is tightly bound to the definition of the space. Therefore, we need to understand how one can give a single characterization that is general enough to encompass the three cases, while constrained enough to recover the correct expression in each case. Fortunately, in our preliminary work we have found the right approach. In all cases, the entropy possesses an upper bound to its increase under convex combination: $S(p\ens[a] + \bar{p} \ens[b]) \leq I(p) + p S(\ens[a]) + \bar{p} S(\ens[b])$ where $\bar{p} = 1-p$ and $I(p) = -p \log p - \bar{p}\log \bar{p}$. This maximum increase is given, in the classical case, when the two measures have disjoint support; in the quantum case, it is given when the eigenspaces of non-zero eigenvalues are orthogonal. Our approach, then, is to proceed in the opposite way: impose a generic entropy function to have an upper bound and enough properties to show that said upper bound must correspond to the Shannon entropy formula~\cite{Shannon}. Since in both classical and quantum mechanics ensembles can always be decomposed into orthogonal components, the different entropy formulas can be recovered in each case as a recursive expansion of the upper bound.

@@

The second axiom imposes a convex structure~\cite{Gudder1973,fritz2015convexspaces}, which represents the ability to perform statistical mixtures. Given two ensembles $\ens[a], \ens[b] \in \Ens$, we can always choose a probability $p \in [0,1]$ and select $\ens[a]$ $p$ percent of the times and $\ens[b]$ $\bar{p} = 1-p$ percent of the times. This gives the ensemble space a convex structure as $p \ens[a] + \bar{p} \ens[b] \in \Ens$ for all $\ens[a], \ens[b] \in \Ens$ and $p \in [0,1]$. This insight is already used in the literature, particularly in Generalized Probabilistic Theories (GPTs); see e.g.~Refs.~\cite{plavala2023,wilce2025generalizedprobabilitytheorynotes,DArianoBook}. The main innovation in this respect is that we do not limit ourselves to finite combinations and we do not simply allow all infinite combinations (which can be shown to include physically untenable cases~\cite{baccetti2013infinite,aop-unphysHilbert}). Rather, we let the topology do the work. That is, infinite convex combinations exist only if they are the topological limit of finite convex combinations. On physical grounds, the mixing operation must be continuous in the natural topology defined by experimental verifiability, much like addition and scalar multiplication are continuous in a topological vector space. Given the analogy, we call \textbf{topological convex space} a convex space that is also a topological space for which the mixing operation is continuous.

Topological convex spaces have not been developed in the literature, and this structure alone presents interesting sets of questions.  For example, it is established that a convex space embeds into a vector space if and only if it is cancellative~\cite{CapraroFritz2013}, that is, if $p \ens[a] + \bar{p} \ens = p \ens[b] + \bar{p} \ens$ for some $p \in (0,1)$ implies $\ens[a] = \ens[b]$. The question now is what are necessary and sufficient conditions for a topological convex space to embed continuously into a topological vector space? Can we have, for example, a Hausdorff second-countable topological convex space that is not a continuous embedding of a topological vector space? Topological vector spaces present a number of nice results which may or may not hold in topological convex spaces. For example, any $\mathsf{T}_1$ second-countable topological vector space is metrizable. Does that same result hold for topological convex spaces? Another interesting aspect is the notion of closure. In a topological vector space, the topological closure of a convex set is convex. We found that this is true in topological convex spaces as well. Additionally, there is another type of closure that may be more interesting physically. We can define the $\sigma$-hull of a set $A$ as the set of possible infinite convex combinations, which can also be shown to be a convex set. This will be a superset of the hull of $A$, because it will include all finite convex combinations, and it will be a subset of the closed hull of $A$, as all infinite convex combinations are topological limits. Moreover, it can be shown that, in some cases, the hull, the $\sigma$-hull and the closed hull are all different. From the physics side, the $\sigma$-hull may be a more physically meaningful closure as it represents the true limit process of statistical mixing, the operation the axiom is characterizing. The subtle issue is that convex combinations always return elements ``inside'' the set, while linear combinations can also return the ``edges.'' Therefore, there are already interesting topics to study even without imposing the entropic structure.

@@

Though we do not believe the current version of the axiom lists all necessary properties of the entropic structure, as discussed below, it does list enough of them to already prove interesting results. For example, we can show that $I(p)$, the maximum entropy increase during mixing, must be, up to a multiplicative constant, the Shannon entropy~\cite{Shannon} $I(p)= - p \log p - \bar{p} \log \bar{p}$. The proof is general in the sense that it does not assume whether the specific model is classical, quantum or something else entirely. Moreover, in the case of both classical and quantum mechanics, states can be decomposed into orthogonal components, which means the upper variability is an equality and can be used to calculate the entropy from that decomposition. This will allow us to recover the expressions for the entropy in the three different cases, classical discrete, classical continuous, and quantum. Similarly, we will be able to show that, from this structure alone, the geometric structures of both classical and quantum mechanics can be recovered.

Our preliminary work also shows that an entropy satisfying those properties forces the convex space to be cancellative, meaning that it embeds into a vector space~\cite{CapraroFritz2013}. That is, a convex space that does not embed into a vector space cannot possess an entropic structure. We can, in fact, produce a model of that case and see exactly why the entropic structure fails. We can also show that an ensemble space must be ``directionally bounded,'' meaning its intersection with any affine line is a finite interval. It is an open question as to whether the ensemble space must be a bounded set. We believe we can also show that some examples of Generalized Probabilistic Theories are inconsistent with an entropic structure and therefore cannot represent a physically meaningful space of ensembles. In other words, the entropic structure constrains the types of spaces possible for physical theories in surprising and interesting ways.

@@

Given an internal point $\ens[a] \in \Ens$ (i.e.~for every $\ens[b] \in \Ens$ we can find $\ens[c] \in \Ens$ such that $\ens[a] = p \ens[b] + \bar{p} \ens[c]$ from some $p \in (0,1)$), we can construct an embedding vector space $V_{\ens[a]}$. If $V_{\ens[a]}$ is finite dimensional, we can show that $V_{\ens[a]}$ is a neighborhood of $\ens[a]$, which means that the interior of $\Ens$ is a manifold. If the entropy is twice differentiable with respect to the convex coefficients, the strict concavity of the entropy means that the Hessian is negative definite, which allows us to define a metric tensor. On a linear coordinate chart we can define $g_{\ens}( v, w ) = - \frac{\partial^2 S}{\partial \ens ^2}( v, w )$ where $v$ and $w$ are elements of the tangent space at $\ens \in \Ens \subset V_{\ens[a]}$. Since all Hessians are symmetric, $g_{\ens}$ will be bilinear, symmetric and positive definite: it is a metric tensor and $V_{\ens[a]}$ is a Hessian manifold~\cite{hessianmanifoldsShima1997,hessianmanifoldsliu2025}. In classical and quantum mechanics it recovers the Fisher information metric (see e.g.~Refs.~\cite{facchi2010,AmariInfoGeom}). The immediate goal, then, is to see how much of information geometry can be generalized to this case. If a locally convex topology can be established, the further goal is to extend this to the infinite-dimensional case.

Another construction allows us to generalize the notion of count of states, given in classical mechanics by the Liouville measure. This can be done by noting two things: for a uniform distribution over the state space $X$ with support $U \subseteq X$, $\mu(U) = 2^{S(\rho_U)}$, and the uniform distribution over $U$ has the highest entropy of all distributions with support $U$. Given a set $A \subseteq \Ens$ of ensembles, we can then define the \textbf{state capacity} as the supremum of the exponential of the entropy within the hull of $A$, which physically means the highest entropy achievable with a statistical mixture of elements of $A$. This set function has nice properties: it is non-negative, monotonic, subadditive in general and additive on orthogonal sets of ensembles. Note that if $A_U \subseteq \Ens$ is the subspace of all probability measures with support $U \subseteq X$, the state capacity of $A_U$ recovers the Liouville measure of $U$. Moreover, $A_U$ is orthogonal to $A_V$ if and only if $U$ and $V$ are disjoint. This means that the state capacity restricted to the lattice of those subspaces is an additive measure. Interestingly enough, one can follow the same construction in the quantum case, and the measure will return the dimensionality of the subspace. Concerning non-additive measures, there are additional interesting questions as one can create a non-additive version of probability measures, which may point to a non-additive version of probability theory. We have already looked at this problem, and, unfortunately, the standard techniques that have been developed in that area are not sufficient to handle the cases of interest to us within physical theories.

@@

The immediate significance of the proposed research stems from the realization that, though classical information theory and quantum information theory are presented as two separate subjects in textbooks, their numerous corresponding features make them an interesting target for generalization.  The broader significance lies in the overall approach of the research as well as its specific potential to provide a common axiomatic foundation for \emph{all} physical theories.  Currently within physics, we typically do not know from where the various mathematical structures emerge, making it unclear in which directions they can be generalized in order to develop new theories that are physically meaningful.  Our approach instead seeks to \emph{find axioms from physical requirements}, i.e.~the axiomatics are justified from the physics.  Such an axiomatic foundation common to all physical theories would serve not only to better ground current theories and elucidate their relationships, but also to constrain potential future physical theories. While there exists other work attempting to reconstruct quantum mechanics from physical principles, for example in Refs.~\cite{hardy2001quantumaxioms,caticha2008,dakic2009,Chiribella2010,Masanes_2011}, we are not aware of any literature attempting to relate the geometric and entropic structures of either classical or quantum mechanics, much less those of a general physical theory. Works trying to reconstruct quantum mechanics from physical premises are typically in finite dimensions, do not generalize to classical mechanics, and do not provide a mathematically rigorous path to generalize to classical or quantum field theories.

@@

\noindent
\textbf{Relationship to other current funding:} PI Aidala's current funding from the NSF Division of Physics is for work on the LHCb experiment at CERN, and from the NSF AccelNet program is to increase collaboration on quantum chromodynamics (QCD) research between the U.S.~and Latin America.  The PI's funding from the Department of Energy is for work at the Relativistic Heavy Ion Collider (RHIC) on the PHENIX and sPHENIX experiments.  These grants are for a completely separate line of research. Her grant from the John Templeton Foundation, titled ``Why classical probability and classical information theory are incompatible with quantum mechanics and quantum contextuality,'' supports the development of novel non-additive measures (set functions) suitable for quantifying the number of states in quantum mechanics~\cite{aop-nonaddmeas}.  There is no direct overlap with the activities proposed here.  Co-PI Uribe does not have any current funding.

### 2025-11-09 Proposals/2025-NSF-EntropicGeometry/ProjectSummary.tex
Fix typos, add refs, remove blue text

\noindent
\textbf{Overview:}\\
\textcolor{blue}{The goal of the proposed work is to develop a common axiomatic basis for the information-theoretic structure underlying all physical theories, classical and quantum. The immediate aim is to generalize results of both classical and quantum information theory to a common, more abstract theory that can also handle the infinite-dimensional case. It takes the notion of statistical ensemble as a fundamental concept, as any physical theory must \emph{at least} provide us with a space of statistical ensembles. The relationships among the four mathematical structures required by a space of ensembles---the topology, the convex structure, the proposed entropic structure and the Lie algebra---will be investigated.} \\

\noindent
\textbf{Intellectual Merit:}  \\
\textcolor{blue}{While the state spaces of classical mechanics and quantum mechanics have different structures, symplectic manifolds and projective spaces of complex Hilbert spaces, respectively, the spaces of classical and quantum \emph{statistical ensembles} have common features.  They are both convex subsets of a locally convex topological vector space whose metric is induced by the information entropy, or more specifically, by the Jensen-Shannon divergence.  Moreover, classical and quantum information theory have numerous corresponding features.  The primary advancement of the research will be to take those common and corresponding features and generalize them, providing a common, mathematically rigorous, foundational structure to both classical and quantum theories.  Such a generalization will provide further insight into the relationships among established physical theories and place constraints on the allowed mathematical structures in potential future physical theories. All axioms will be physically motivated and stated as properties of ensemble spaces.  Preliminary work shows that the symplectic form of classical mechanics and the inner product (i.e.~Born rule) of quantum mechanics can be recovered by the entropy, tightly linking the geometric and entropic structures, and that the states defined by mechanics can be understood as special cases of ensembles with the lowest possible entropy or the extreme points of the convex set. }

>>>>

\noindent
\textbf{Overview:}\\
The goal of the proposed work is to develop a common axiomatic basis for the information-theoretic structure underlying all physical theories, classical and quantum. The immediate aim is to generalize results of both classical and quantum information theory to a common, more abstract theory that can also handle the infinite-dimensional case. It takes the notion of statistical ensemble as a fundamental concept, as any physical theory must \emph{at least} provide us with a space of statistical ensembles. The relationships among the four mathematical structures required by a space of ensembles---the topology, the convex structure, the proposed entropic structure and the Lie algebra---will be investigated. \\

\noindent
\textbf{Intellectual Merit:}  \\
While the state spaces of classical mechanics and quantum mechanics have different structures, symplectic manifolds and projective spaces of complex Hilbert spaces, respectively, the spaces of classical and quantum \emph{statistical ensembles} have common features.  They are both convex subsets of a locally convex topological vector space whose metric is induced by the information entropy, or more specifically, by the Jensen-Shannon divergence.  Moreover, classical and quantum information theory have numerous corresponding features.  The primary advancement of the research will be to take those common and corresponding features and generalize them, providing a common, mathematically rigorous, foundational structure to both classical and quantum theories.  Such a generalization will provide further insight into the relationships among established physical theories and place constraints on the allowed mathematical structures in potential future physical theories. All axioms will be physically motivated and stated as properties of ensemble spaces.  Preliminary work shows that the symplectic form of classical mechanics and the inner product (i.e.~Born rule) of quantum mechanics can be recovered by the entropy, tightly linking the geometric and entropic structures, and that the states defined by mechanics can be understood as special cases of ensembles with the lowest possible entropy or the extreme points of the convex set.
